---
title: "Project - California Quail in BC"
author: "Tyler Stevenson, Foster Lockerbie, Seamus Riordan-Short"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  #pdf_document: default
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Loading Required Packages

```{r, message = FALSE}
#load packages
library(spatstat)
library(sf)
library(rgbif)
```

## Import Data

```{r, cache=TRUE}
#occ_download_wait('0002650-250402121839773')
data <- occ_download_get('0002650-250402121839773') %>%
  occ_download_import()
```

```{r, cache=TRUE}
# Filter for only British Columbia records
bc_quail <- data[data$stateProvince %in% 
                             c("British Columbia", 
                               "British Columbia (BC)", 
                               "British Columbia (Prov.)",
                               "Bc"), ]

```

## Convert Data to PPP

```{r, cache=TRUE}
load("BC_Covariates.Rda")

#Identify Class of Each Element
sapply(DATA, class)

# Convert quail data to sf object
quail_df <- st_as_sf(bc_quail, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE)

# Remove duplicated coordinates
quail_df <- quail_df[!duplicated(st_coordinates(quail_df)), ]

# Convert window to sf
window_sf <- st_as_sf(DATA$Window)

# Ensure both quail coordinates and the window are in the same CRS
quail_df <- st_transform(quail_df, crs = st_crs(window_sf))

#Remove Points Outside the Window
inside_window <- st_intersects(quail_df, window_sf)
quail_df <- quail_df[sapply(inside_window, length) > 0, ]

# fill na with 1
# quail_df$individualCount <- tidyr::replace_na(quail_df$individualCount, 1)

#use uncount to unpack multiple observations
# quail_df <- quail_df %>% tidyr::uncount(weights = individualCount)

# Extract the x and y coordinates of the quail
quail_df <- st_as_sf(quail_df, coords = c("decimalLongitude", "decimalLatitude"), crs = st_crs(window_sf), remove = FALSE)
quail_coords <- st_coordinates(quail_df)

# Create the ppp object
quail_ppp <- ppp(x = quail_coords[, 1], y = quail_coords[, 2], window = as.owin(window_sf))
```

```{r}
par(mar = c(0, 0, 0, 0), bg = "#ACB2BE")
plot(quail_ppp, main = "", pch = 16, cex = 0.4, col = "#543111", cols = adjustcolor("#EFD8B6", alpha.f = 0.3), background = "red")
```

## Determine Homogeneity

```{r, cache=TRUE}
# Split into quadrat and count points
Q <- quadratcount(quail_ppp, nx = 5, ny = 5)
```

```{r}
#Plot the quadrats and points
par(mar = c(0, 0, 0, 0), bg = "#ACB2BE")
plot(quail_ppp, pch=16,cex=.4, use.marks = F, col = "#543111",
     cols= adjustcolor("#EFD8B6", alpha.f = 0.5), main = "")
plot(Q, cex=2, col="#8C725A", add=T)
```


```{r}
#Plot the intensity
plot(intensity(Q, image =T), main="Intensity")
plot(quail_ppp, pch=16, cex=.5, use.marks = F, cols= "white",add=T)
```

```{r}
#Quadrat test of homogeneity 
quadrat.test(Q)
```

The small pvalue suggests that there is significant deviation from homogeneity. The quails tend to be clustered in some areas and others have none. The intensity varies in each quadrat suggesting that the intensity is not homogeneous and is spatially varying.

## Plotting Intensity

```{r, cache=TRUE}
#kernel estimation with likelihood cross validation bandwidth selection
lambda_u_hat <- density(quail_ppp, sigma = bw.ppl)
plot(lambda_u_hat, main="Hotspots")

# Estimate R
R <- bw.ppl(quail_ppp)

#Calculate test statistic
LR <- scanLRTS(quail_ppp, r = R)

#Plot the output 
plot(LR)
plot(as.owin(window_sf),add=T,border="green")

#Compute local p-values
pvals <- eval.im(pchisq(LR,
                        df = 1,
                        lower.tail = FALSE))


#Plot the output
plot(pvals, main = "Local p-values")
plot(as.owin(window_sf),add=T)
```

```{r, cache=TRUE}
#kernel estimation with likelihood cross validation bandwidth selection
par(mar = c(0, 0, 0, 0), bg = "#ACB2BE")
lambda_u_hat <- density(quail_ppp, sigma = bw.ppl, main = "")
plot(lambda_u_hat, main="", box = FALSE, col = viridis::inferno)
```


The quail appear to be located in the southern interior and the southern coast of BC.

## Relationship Between Quail

```{r, cache=TRUE}
#Estimate a strictly positive density
lambda_pos <- density(quail_ppp, sigma=bw.ppl, positive=TRUE)

#Simulation envelope (with points drawn from the estimated intensity)
E_inhom <- envelope(quail_ppp,
                    Kinhom,
                    simulate = expression(rpoispp(lambda_pos)),
                    correction="border",
                    rank = 1,
                    nsim = 19,
                    fix.n = TRUE)
# visualise the results
plot(E_inhom,
     main = "",
     lwd = 2)

# Estimate g corrected for inhomogeneity
#g_inhom <- pcfinhom(quail_ppp)

#Simulation envelope (with points drawn from the estimated intensity)
#pcf_inhom <- envelope(quail_ppp,
                      #pcfinhom,
                      #simulate = expression(rpoispp(lambda_pos)),
                      #rank = 1,
                      #nsim = 19)

# visualise the results
#plot(pcf_inhom)
plot(E_inhom,
     main = "",
     lwd = 2,
     xlim = c(0, 50000))
```

There appears to be clustering at lower distances. Which makes sense since quail tend to live in groups called coveys.

## Developing Model

### Covariates

```{r}
#Estimate Rho for elevation
rho_elev <- rhohat(quail_ppp, DATA$Elevation)

plot(rho_elev,
     main = "",
     xlab = "Elevation (m)",
     xlim = c(0, 3000))

```

```{r}
plot(rho_elev,
     main = "",
     xlab = "Elevation (m)",
     xlim = c(0, 1000))
```

There appears to be a strong relationship with quail at lower elevations

```{r,cache=TRUE}
#Estimate Rho for forest cover
rho_for <- rhohat(quail_ppp, DATA$Forest)

plot(rho_for,
     main = "",
     xlab = "Forest Cover (%)")
```

There also appears to be a relationship with lower % of forest cover

```{r,cache=TRUE}
#Estimate Rho for HFI
rho_hfi <- rhohat(quail_ppp, DATA$HFI)

plot(rho_hfi,
     main = "",
     xlab = "HFI")
```

There appears to be a relationship with higher levels of Human Footprint Index

```{r,cache=TRUE}
#Estimate Rho for distance to water
rho_wat <- rhohat(quail_ppp, DATA$Dist_Water)

plot(rho_wat,
     main = "",
     xlab = "Distance to Water")
```

There appears to be a relationship with distance to water. The quail appear to be more present very close to the water.

### Collinearity

```{r,cache=TRUE}
#Check for collinearity
cor.im(DATA$Elevation, DATA$Forest, DATA$HFI, DATA$Dist_Water, use = "complete.obs")
```

There does not appear to be collinearity between the covariates

### Initial Model

Elevation, Forest Cover, HFI all appear to have a non-linear relationship. Distance to Water appears very sporadic but may be captured with linear

```{r,cache=TRUE}
# initial PPM model
fit <- ppm(quail_ppp ~ Elevation + I(Elevation^2) + Forest + I(Forest^2) + Dist_Water + HFI + I(HFI^2), data = DATA)

fit

#Fit the null model
fit_null <- ppm(quail_ppp ~ 1)

fit_null

#AIC values
AIC(fit); AIC(fit_null)

#Delta AIC
AIC(fit_null) - AIC(fit)

#Likelihood ratio test
anova(fit_null, fit, test = "LRT")
```

The AIC is lower and the likelihood ratio test is significant so the model we defined is a better fit than the null model.

```{r,cache=TRUE}
#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit, "Elevation")

#Calculate the partial residuals as a function of forest cover
par_res_for <- parres(fit, "Forest")

#Calculate the partial residuals as a function of HFI
par_res_hfi <- parres(fit, "HFI")

#Calculate the partial residuals as a function of forest cover
par_res_wat <- parres(fit, "Dist_Water")

#Side by side plotting
par(mfrow = c(2,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_for,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Cover (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI")
plot(par_res_wat,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
```

```{r}
#Plot the model predictions
par(mar = c(0, 0, 0, 0), bg = "#ACB2BE")
plot(fit,
     se = FALSE,
     box = FALSE,
     log = TRUE,
     superimpose = FALSE,
     main = "",
     col = viridis::inferno,
     n= 256)

```

### Model Adjustments with GAMS

```{r}
library(splines)

#Fit the PPP model
fit_smooth <- ppm(quail_ppp ~ bs(Elevation,8) + bs(Forest, 8) + bs(HFI, 8)+ Dist_Water, data = DATA, use.gam = TRUE)

fit_smooth

#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit_smooth, "Elevation")

#Calculate the partial residuals as a function of forest cover
par_res_for <- parres(fit_smooth, "Forest")

#Calculate the partial residuals as a function of HFI
par_res_hfi <- parres(fit_smooth, "HFI")

#Calculate the partial residuals as a function of forest cover
par_res_wat <- parres(fit_smooth, "Dist_Water")

#Side by side plotting
par(mfrow = c(2,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_for,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Cover (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI")
plot(par_res_wat,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
```

```{r}
#AIC values
AIC(fit); AIC(fit_smooth)

#Delta AIC
AIC(fit) - AIC(fit_smooth)

#Likelihood ratio test
anova(fit, fit_smooth, test = "LRT")
```

```{r}
#Plot the model predictions
plot(fit_smooth,
     se = FALSE,
     log = TRUE,
     superimpose = FALSE,
     main = "Estimated Quail intensity",
     n= 256)

```

### Model Adjustments with Cartesian Coordinates

```{r}
fit_cart <- ppm(quail_ppp ~ Elevation + I(Elevation^2) + Forest + I(Forest^2) + Dist_Water + HFI + I(HFI^2)+y, data = DATA)

fit_cart

#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit_cart, "Elevation")

#Calculate the partial residuals as a function of forest cover
par_res_for <- parres(fit_cart, "Forest")

#Calculate the partial residuals as a function of HFI
par_res_hfi <- parres(fit_cart, "HFI")

#Calculate the partial residuals as a function of forest cover
par_res_wat <- parres(fit_cart, "Dist_Water")

#Calculate the partial residuals as a function of y
par_res_y <- parres(fit_cart, "y")

#Side by side plotting
par(mfrow = c(3,2))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_for,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest Cover (%)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI")
plot(par_res_wat,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
plot(par_res_y,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Cartesian y")
```

```{r}
#AIC values
AIC(fit); AIC(fit_cart)

#Delta AIC
AIC(fit) - AIC(fit_cart)

#Likelihood ratio test
anova(fit, fit_cart, test = "LRT")
```

```{r}
#Plot the model predictions
par(mar = c(0, 0, 0, 0), bg = "#ACB2BE")
plot(fit_cart,
     se = FALSE,
     box = FALSE,
     log = TRUE,
     superimpose = FALSE,
     main = "",
     col = viridis::inferno,
     n= 256)
```
